Dynamic Species Interactions in the San Nicolas Island Kelp Forest

### 2018-07-19 EDM_SNI_analysis.Rmd
ported manuscript to markdown

```{r setup, include=FALSE,message=F}
# Load required packages
require(tidyverse)
require(rEDM)
require(quantreg)
require(knitr)
require(kableExtra)
require(extrafont)
require(GGally)
require(sna)
require(network)

@@

```{r network,fig.height=6,fig.width=6}
connect.mat <- t(westend.ptot.mat) %>% 
  as.data.frame() %>%
  mutate(mei=rep(0,nrow(.)),pdo=rep(0,nrow(.)),npgo=rep(0,nrow(.)),waves=rep(0,nrow(.)),sst=rep(0,nrow(.)))%>%
  as.matrix()
names <- map_chr(colnames(connect.mat), function(x) as.character(namekey$plotting[match(x,namekey$short)]))
names [7:11] <- c("MEI","PDO","NPGO","Max Hs","SST")
g <- network(connect.mat,vertex.attrnames = names,loops=F)
g %v% "type" <- c(rep("darkgreen",3),rep("darkorchid1",2),"darkgreen",rep("dodgerblue",5))
g %v% "labels" <- names
int_net<-ggnet2(g,arrow.size=8,arrow.gap=0.05,color = "type")+
  geom_point(aes(color = color), size = 12, alpha = 0.5) +
  geom_point(aes(color = color), size = 9)+
  geom_text(aes(label=label),family="Rockwell",size=4)+
  theme(panel.border = element_rect(colour="white",fill=NA))

ggsave("plots/int_net.png",plot=int_net)
int_net
```

@@

```{r 3d plot,fig.height=7,fig.width=7}
require(plot3D)

@@

scatter3D(x[110],y[110],z[110],col="blue",pch=19,cex=2,add=T)
scatter3D(x[236],y[236],z[236],col="blue",pch=19,cex=2,add=T)
scatter3D(x[170],y[170],z[170],col="blue",pch=19,cex=2,add=T)

dev.copy(pdf,"plots/multi_attractor.pdf")
dev.off()

par(op)
rm(xsmooth,ysmooth,zsmooth,xtemp,ytemp,ztemp,t,tt,seg.start,seg.end,x,y,z,temp)

@@

```{r individual_effects,fig.width=7}
phys_effects_on_interactions <- purrr::map(as.list(study_spp),function(spp) {
  quants <- c(0.05,0.5,0.95)
  plot_title <- namekey$long[match(spp,namekey$short)]
  spp.only.coeffs.with.phys %>%
  filter(!is.na(type),mod==spp)%>%
  ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
    # geom_linerange(aes(ymin=lower,ymax=upper),col='gray60')+
    geom_point(col='gray60',alpha=0.6)+
    geom_hline(yintercept = 0,linetype=2)+
    geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
    scale_color_manual(values=c('mac'="navyblue",'ymac'="gray30",'pter'="darkgreen",'lam'="darkcyan",'purp'="mediumpurple4",'red'="darkred"),name="effect")+
    labs(title=paste("Physical forcing effect on interactions with",plot_title),x="Normalized Index Value",y="Interaction Strength")+
    facet_grid(type~external_forcing,scales = "free_y")+
    theme(strip.text=element_text(size=8),
          legend.text = element_text(size=12),
          legend.position="right",
          legend.margin=margin(t = -0.5, unit='line'),
          plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line")
        )
})
ggsave("plots/phys_effects_red.png",plot=phys_effects_on_interactions[[1]])
ggsave("plots/phys_effects_purp.png",plot=phys_effects_on_interactions[[2]])
ggsave("plots/phys_effects_lam.png",plot=phys_effects_on_interactions[[3]])
ggsave("plots/phys_effects_pter.png",plot=phys_effects_on_interactions[[4]])
ggsave("plots/phys_effects_mac.png",plot=phys_effects_on_interactions[[5]])
ggsave("plots/phys_effects_ymac.png",plot=phys_effects_on_interactions[[6]])

@@

```{r mac pter}
mac_effect_on_pter <-spp.only.coeffs.with.phys %>%
  ungroup()%>%
  filter(mod=="pter",effect=="mac")%>%
  mutate(phys_high_low=case_when(
    ext_value > 1 ~ "high",
    ext_value < -1 ~ "low")) %>%
  filter(!is.na(phys_high_low))%>%
  left_join(select(westend,period,one_of(phys.vars)),by="period")%>%
  group_by(type,external_forcing,phys_high_low)%>%
  mutate(mean_int=mean(value,na.rm=T))%>%
  distinct()%>%
  ungroup()

mac_effect_on_pter %>%
ggplot(aes(value,..density..,fill=phys_high_low))+
  geom_density(col=NA,alpha=0.6)+
  geom_vline(xintercept=0,linetype=2)+
  geom_vline(aes(xintercept=mean_int,color=phys_high_low))+
  scale_fill_manual(values=c("red","blue"),name="Physical\nForcing Value")+
  scale_color_manual(values=c("red","blue"))+
  guides(color="none")+
  labs(x="Effect of Macrocystis on Pterygophora",y="Probability Density")+
  facet_grid(rows=vars(external_forcing))

@@

```{r interactions density plot by type,fig.height=6}
# Density plot
interaction_distribution_by_type <- spp.only.mod.coeffs %>%
  filter(!is.na(type))%>%
  group_by(type) %>%
  mutate(mean_int=mean(value,na.rm=T))%>%
  ungroup()%>%
  ggplot(aes(value,..density..))+
  geom_density(aes(fill=type),alpha=0.8,col=NA)+
  geom_vline(xintercept=0,linetype=2)+
  geom_vline(aes(xintercept=mean_int))+
  scale_fill_manual(values=c("darkorange1","orangered2","darkorchid1","palegreen3","skyblue1"),name="")+
   guides(fill=F)+
  xlim(-2,2)+
  labs(x="Interaction strength",y="Kernel density")+
  geom_text(aes(x=-2,y=0.5,label=type),hjust="inward",family="Rockwell",fontface="italic",check_overlap = TRUE,size=4,show.legend = FALSE)+
  facet_grid(rows=vars(type))+
  theme(strip.text = element_blank())

interaction_distribution_by_type
ggsave("plots/int_dist_by_type_density.png",plot=interaction_distribution_by_type,h=6)
```

@@

```{r mean vs variance}
int_mean_sd <-spp.only.mod.coeffs %>%
  filter(!is.na(type))%>%
  mutate(neutral=ifelse(lower<0 & upper>0,1,0))%>%
  group_by(mod,effect,type) %>%
  summarise(mean_int=mean(value,na.rm=T),abs_mean=abs(mean_int),sd_int=sd(value,na.rm=T),n=n(),cv_int=sd_int/abs_mean*100,
            perc_neutral=sum(neutral)/n()*100)%>%
  ungroup()

# Mean interaction strength by interaction type
int_mean_sd %>%
  filter(mod != "ymac",effect != 'ymac')%>%
  group_by(type)%>%
  summarise(mean_str=mean(abs_mean),ci95=sd(abs_mean)/sqrt(n())*1.96)%>%
  ungroup()%>%
  ggplot(aes(type,mean_str))+
  geom_col()+
  geom_errorbar(aes(ymax=mean_str+ci95,ymin=mean_str-ci95),position='dodge',width=0.2)+
  labs(x='Interaction Type',y='Interaction Strength')

# scatterplot mean vs. sd (removing intraspecies interactions)
int_mean_sd %>%
  filter(mod != "ymac",effect != 'ymac')%>%
  ggplot(aes(abs_mean,sd_int))+
  geom_point(aes(color=type),size=5)+
  scale_color_manual(values=c("darkorange1","orangered2","darkorchid1","palegreen3","skyblue1"),name="")+
  geom_smooth(method="lm",se=F,col="black")+
  labs(x="Mean Interaction Strength",y="SD Interaction Strength")

# CV calculated using the correction for sample size (cv*=1+1/4n * cv)
int_mean_sd %>%
  filter(mod != 'ymac',effect!='ymac')%>%
  ggplot(aes(type,cv_int))+
  geom_col()+

  labs(x="Interaction Type",y="CV of Measured Interactions")

@@

### Bidirectional interactions (is the effect of A on B the same as B on A)
```{r bidirectional}
int_mean_sd %>%
  mutate(rev_effect=mod,rev_mod=effect) %>%
  left_join(int_mean_sd,by=c("rev_effect"="effect","rev_mod"="mod"))%>%
  mutate(upper1=mean_int.x+sd_int.x,lower1=mean_int.x-sd_int.x,upper2=mean_int.y+sd_int.y,lower2=mean_int.y-sd_int.y)-> int_mean_sd2

int_mean_sd2 %>%
  filter(type.x %in% c('algal competition','urchin competition'),type.y %in% c('algal competition','urchin competition'))%>%
  ggplot(aes(mean_int.x,mean_int.y))+
  geom_point()+
  geom_smooth(method="lm")+
  geom_abline(slope=1,intercept=0,linetype=2)+
  geom_errorbar(aes(ymin=lower2,ymax=upper2))+
  geom_errorbarh(aes(xmin=lower1,xmax=upper1))+
  labs(x='Effect of A on B',y='Effect of B on A')
```

>>>>

```{r setup, include=FALSE,message=F}
# Load required packages
require(tidyverse)
require(rEDM)
require(quantreg)
require(knitr)
require(kableExtra)
require(extrafont)
require(GGally)
require(ggalt)
require(sna)
require(network)

@@

```{r network,fig.height=6,fig.width=6}
# g <- network(connect.mat,vertex.attrnames = names,loops=F)
# g %v% "type" <- c(rep("darkgreen",3),rep("darkorchid1",2),"darkgreen",rep("dodgerblue",5))
# g %v% "labels" <- names
# set.edge.value(g,"weights",edge.weight.mat*3)
# 
# int_net<-ggnet2(g,arrow.size=6,arrow.gap=0.04,color = "type",label=names,label.alpha=0,edge.size="weights")+
#   geom_point(aes(color = color), size = 14, alpha = 0.5) +
#   geom_point(aes(color = color), size = 12)+
#   geom_text(aes(label=label),colour='white',family="Rockwell",size=3)+
#   theme(panel.border = element_rect(colour="white",fill=NA))
# 
# # ggsave("plots/int_net.pdf",plot=int_net)
# int_net

## i graph
require(igraph)

# connections matrix for the network based on the CCM results above
connect.mat <- t(westend.ptot.mat) %>% 
  as.data.frame() %>%
  mutate(mei=rep(0,nrow(.)),pdo=rep(0,nrow(.)),npgo=rep(0,nrow(.)),waves=rep(0,nrow(.)),sst=rep(0,nrow(.)))%>%
  as.matrix()

# edges (arrows) weighted based on CCM rho value
edge.weight.mat <- t(westend.xmap_mat)%>%
  as.data.frame() %>%
  mutate(mei=rep(0,nrow(.)),pdo=rep(0,nrow(.)),npgo=rep(0,nrow(.)),waves=rep(0,nrow(.)),sst=rep(0,nrow(.)))%>%
  as.matrix()
edge.weight.mat[is.na(edge.weight.mat)]<-0
edge.weight.mat

names <- map_chr(colnames(connect.mat), function(x) as.character(namekey$plotting[match(x,namekey$short)]))
names[6] <- "M. pyr\n(j)"
names [7:11] <- c("MEI","PDO","NPGO","Max Hs","SST")

# initiate the igraph
g2 <- graph.adjacency(connect.mat)

# Vertex labels
V(g2)$name <-names

# label with the number of incoming(outgoing) connections
vert_in <- sapply(igraph::degree(g2,mode="in"),function(x) paste0("(",x,")"))
vert_out_in <- paste0(igraph::degree(g2,mode="out"),vert_in)
V(g2)$label <- paste(sep="\n",V(g2)$name, vert_out_in)
V(g2)$label[7:11] <- paste(V(g2)$name[7:11])

# #vertex coordinates (layout of plot
# crds <- cbind(c(116.0000, 318.0000, 177.0000, 581.0000, 346.0000, 385.0000, 331.1627, 151.8379, 573.0000, 539.0000,  45.0000),c(0.0000, 186.0000, 362.0000, 172.0000,  40.0000, 374.0000, 537.0000, 504.2360, 35.0000, 406.0000, 288.4111))

# vector of node colors for plotting
vert.cols <- c(rep("#238B45",3),rep("#BEBADA",2),"#238B45",rep("#8DD3C7",5))

# weighted edges, with width and color
edge.weights <- as.numeric(t(edge.weight.mat))[t(edge.weight.mat) != 0]*5
E(g2)$width <- edge.weights
col.1 <- adjustcolor("#99D8C9", alpha=0.8)
col.2 <- adjustcolor("#006D2C", alpha=0.8)
edge.pal <- colorRampPalette(c(col.1, col.2), alpha = TRUE)
edge.pal <- edge.pal(100)
edge.cols<-round((edge.weights-min(edge.weights))/(max(edge.weights)-min(edge.weights))*100) +1
edge.cols <- edge.pal[edge.cols]

# some edges must be curved to look okay in the graph
curves <- rep(0,40)
curves[c(1,4,13,17,18,21)] <-0.4
curves[c(10,19)] <-0.3
curves[20] <- -0.1
curves[c(6,39)] <- 0.2

#plot to file if uncommented
# png("plots/ccm_network.png",height=600,width=600)
plot(g2,asp=0.8, vertex.label.color="black",vertex.label.family="Times", vertex.label.cex=0.6, edge.arrow.size=0.8,edge.arrow.width=0.8,vertex.label.font=2, vertex.shape="circle", edge.curved=curves,margin=0,vertex.size=20,rescale=T,edge.lty=1,edge.color=edge.cols,layout=layout_in_circle,vertex.color=vert.cols,vertex.frame.color="black",main="")
# dev.off()
```

@@

```{r 3d plot,fig.height=4,fig.width=4,dpi=300}
require(plot3D)

@@

scatter3D(x[110],y[110],z[110],col="blue",pch=19,cex=2,add=T)
scatter3D(x[236],y[236],z[236],col="blue",pch=19,cex=2,add=T)
scatter3D(x[170],y[170],z[170],col="blue",pch=19,cex=2,add=T)
# 
# dev.copy(png,"plots/multi_attractor.png",width=700,height=700,res=96)

par(op)
rm(xsmooth,ysmooth,zsmooth,xtemp,ytemp,ztemp,t,tt,seg.start,seg.end,x,y,z,temp)
```

@@

```{r individual_effects,fig.width=7}
phys_effects_on_interactions <- purrr::map(as.list(study_spp),function(spp) {
  quants <- c(0.05,0.5,0.95)
  plot_title <- namekey$long[match(spp,namekey$short)]
  spp.only.coeffs.with.phys %>%
  filter(!is.na(type),mod==spp)%>%
  ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
    # geom_linerange(aes(ymin=lower,ymax=upper),col='gray60')+
    geom_point(col='gray60',alpha=0.6)+
    geom_hline(yintercept = 0,linetype=2)+
    geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
    scale_color_manual(values=c('mac'="navyblue",'ymac'="gray30",'pter'="darkgreen",'lam'="darkcyan",'purp'="mediumpurple4",'red'="darkred"),name="effect")+
    labs(title=paste("Physical forcing effect on interactions with",plot_title),x="Normalized Index Value",y="Interaction Strength")+
    facet_grid(type~external_forcing,scales = "free_y")+
    theme(strip.text=element_text(size=8),
          legend.text = element_text(size=12),
          legend.position="right",
          legend.margin=margin(t = -0.5, unit='line'),
          plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line")
        )
})
# ggsave("plots/phys_effects_red.png",plot=phys_effects_on_interactions[[1]])
# ggsave("plots/phys_effects_purp.png",plot=phys_effects_on_interactions[[2]])
# ggsave("plots/phys_effects_lam.png",plot=phys_effects_on_interactions[[3]])
# ggsave("plots/phys_effects_pter.png",plot=phys_effects_on_interactions[[4]])
# ggsave("plots/phys_effects_mac.png",plot=phys_effects_on_interactions[[5]])
# ggsave("plots/phys_effects_ymac.png",plot=phys_effects_on_interactions[[6]])

@@

```{r mac pter,fig.width=6,fig.height=6}
mac_effect_on_pter <-spp.only.coeffs.with.phys %>%
  ungroup()%>%
  filter(mod=="pter",effect=="mac")%>%
  mutate(phys_high_low=case_when(
    ext_value > 1 ~ "high",
    ext_value < -1 ~ "low")) %>%
  filter(!is.na(phys_high_low))%>%
  left_join(select(westend,period,one_of(phys.vars)),by="period")%>%
  group_by(type,external_forcing,phys_high_low)%>%
  mutate(mean_int=mean(value,na.rm=T))%>%
  distinct()%>%
  ungroup()

# physical forcings labeller
phys.labeller <- c(
  mei= "MEI",
  npgo= "NPGO",
  pdo= "PDO",
  sst="SST",
  waves="SWH"
)

mac_effect_on_pter_plot <-mac_effect_on_pter %>%
ggplot(aes(value,..density..,fill=phys_high_low))+
  geom_density(col=NA,alpha=0.6)+
  geom_vline(xintercept=0,linetype=2)+
  geom_vline(aes(xintercept=mean_int,color=phys_high_low),size=1.5)+
  scale_fill_manual(values=c("darkorange","dodgerblue4"),name="Physical\nForcing Value",labels=c("high","low"))+
  scale_color_manual(values=c("darkorange","dodgerblue4"))+
  guides(color="none")+
  labs(x="Effect of Macrocystis on Pterygophora",y="Probability Density")+
  facet_grid(rows=vars(external_forcing),labeller=labeller(external_forcing=phys.labeller))+
  theme(strip.text=element_text(size=10))

# ggsave("plots/mac_effect_on_pter_phys.png",plot=mac_effect_on_pter_plot,h=6,w=6,dpi=500)
mac_effect_on_pter_plot
```

@@

```{r interactions density plot by type,fig.height=6,fig.width=6}
# Density plot
interaction_distribution_by_type <- spp.only.mod.coeffs %>%
  filter(!is.na(type))%>%
  group_by(type) %>%
  mutate(mean_int=mean(value,na.rm=T))%>%
  ungroup()%>%
  ggplot(aes(value,..density..))+
  geom_density(aes(fill=type),alpha=0.8,col=NA)+
  geom_vline(xintercept=0,linetype=2)+
  geom_vline(aes(xintercept=mean_int))+
  scale_fill_manual(values=c("darkorange1","orangered2","darkorchid1","palegreen3","skyblue1"),name="")+
   guides(fill=F)+
  xlim(-1.5,1.5)+
  labs(x="Interaction strength",y="Kernel density")+
  geom_text(aes(x=-1.5,y=0.5,label=type),hjust="inward",family="Rockwell",fontface="italic",check_overlap = TRUE,size=4,show.legend = FALSE)+
  facet_grid(rows=vars(type))+
  theme(strip.text = element_blank())

interaction_distribution_by_type
# ggsave("plots/int_dist_by_type_density.png",plot=interaction_distribution_by_type,h=6,w=6,dpi=500)
```

@@

```{r mean vs variance}
int_mean_sd <-spp.only.mod.coeffs %>%
  filter(!is.na(type))%>%
  mutate(neutral=ifelse(lower<0 & upper>0,1,0))%>%
  group_by(mod,effect,type) %>%
  summarise(mean_int=mean(value,na.rm=T),abs_mean=abs(mean_int),sd_int=sd(value,na.rm=T),n=n(),
            cv_int=sd_int/abs_mean*100,upper=quantile(value,0.975),lower=quantile(value,0.025),
            perc_neutral=sum(neutral)/n()*100)%>%
  ungroup()

# Mean interaction strength by interaction type
spp.only.mod.coeffs %>%
  filter(!is.na(type))%>%
  mutate(neutral=ifelse(lower<0 & upper>0,1,0))%>%
  filter(mod != "ymac",effect != 'ymac')%>%
  group_by(type) %>%
  summarise(mean_int=mean(value,na.rm=T),abs_mean=abs(mean_int),n=n(),upper=quantile(value,0.975),lower=quantile(value,0.025))%>%
  ungroup()%>%
  ggplot(aes(type,mean_int))+
  geom_col()+
  geom_errorbar(aes(ymax=upper,ymin=lower),position='dodge',width=0.2)+
  labs(x='Interaction Type',y='Interaction Strength')

# scatterplot mean vs. sd (removing intraspecies interactions)
int_mean_sd_plot <- int_mean_sd %>%
  filter(type != "intraspecies")%>%
  ggplot(aes(mean_int,sd_int))+
  geom_point(aes(color=type),size=5)+
  geom_smooth(method="lm",se=F,col="black",show.legend = F)+
  geom_vline(xintercept=0,linetype=2)+
  # geom_encircle(aes(fill=type),col=NA,alpha=0.3,expand=0.1)+
  scale_color_manual(values=c("darkorange1","orangered2","darkorchid1","palegreen3","skyblue1"),name="")+
  scale_fill_manual(values=c("darkorange1","orangered2","darkorchid1","palegreen3","skyblue1"),name="")+
  guides(fill='none')+
  labs(x="Mean Interaction Strength",y="SD Interaction Strength")+
  theme(legend.position = c(0.2,0.8))
# ggsave("plots/int_mean_sd.png",plot=int_mean_sd_plot,dpi=500)
int_mean_sd_plot

# CV
spp.only.mod.coeffs %>%
  filter(!is.na(type))%>%
  filter(type != "intraspecies")%>%
  group_by(type) %>%
  summarise(abs_mean=mean(abs(value)),sd_int=sd(value,na.rm=T),cv_int=abs_mean/sd_int*100)%>%
  ungroup()%>%
  ggplot(aes(type,cv_int))+
  geom_col()+
  labs(x="Interaction Type",y="CV of Measured Interactions")

@@

### Bidirectional interactions (is the effect of A on B the same as B on A)
```{r bidirectional}
int_mean_sd %>%
  mutate(rev_effect=mod,rev_mod=effect) %>%
  left_join(int_mean_sd,by=c("rev_effect"="effect","rev_mod"="mod"))-> int_mean_sd2

int_mean_sd2 %>%
  filter(type.x %in% c('algal competition','urchin competition'),type.y %in% c('algal competition','urchin competition'))%>%
  ggplot(aes(mean_int.x,mean_int.y))+
  geom_point()+
  geom_smooth(method="lm")+
  geom_abline(slope=1,intercept=0,linetype=2)+
  geom_errorbar(aes(ymin=lower.y,ymax=upper.y))+
  geom_errorbarh(aes(xmin=lower.x,xmax=upper.x))+
  labs(x='Effect of A on B',y='Effect of B on A')
```
